A Survey of Possible Exoskeleton Control Architectures and Algorithms\\ (Draft 3.0)

### 2015-09-24 control.tex
cga updates

\newcommand{\gc}{{\mbox {GravityCompensation}}}
\newcommand{\invdyn}{{\mbox {\tiny ID}}}
\newcommand{\invdynw}{{\mbox {\tiny ID+Fw}}}

@@

\title{A Survey of Possible Exoskeleton Control Architectures and
Algorithms\\
(Draft 2.3)}

@@

It is useful to provide human programmers
with idealized abstractions of control of a single degree of
freedom. Here are some common alternatives that could be provided by low level control:
\begin{itemize}
\item
Position control. The joint moves to a specified position and stays there independent
of load.
\item
Velocity control. The joint moves at a specified velocity independent of load.
\item
Force or torque control. The joint provides a specified force or torque independent
of its motion.
\item
Impedance control: The user specifies a desired position (for a spring), and commands
a desired
stiffness, damping, and potentially a modified inertia or moment of inertia.
Modifying inertia is more dangerous than providing variable active stiffness
or damping.
Or (almost equivalently),
the user specifies a desired force or torque
which changes according to the position, velocity,
and acceleration of the joint.
\item
Online optimization of accelerations, actuator forces and torques, and contact forces
using quadratic programming. See Section \ref{sec:qp} for further explanation.
\end{itemize}

@@

PID or more complex feedback control laws are implemented by low level control.

@@

\section{Exoskeleton dynamics}

Exoskeletons are similar to robots, but there are several important differences.
There are two sets of contact forces: operator-exoskeleton (which we denote
with a subscript $o$ as in $\vf_o$) and exoskeleton-world (which we denote
with a subscript $w$ as in $\vf_w$). If the operator is tightly strapped or
rigidly held by other types of physical constraints, we can lump the operator
body parts with the exoskelton parts which they are attached to in terms
of link inertias, locations of center of mass, and moments of inertia,
and just think of the whole system as one robot.
We can also separate out the operator dynamics if this is more convenient
(we separate them out and ignore them in what follows).
If the operator is loosely connected to the exoskeleton, this introduces a
huge modeling and control problem, and would also be very dangerous for the
operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).
We do not address the loosely connected case in this paper.

The dynamics of a revolute exoskeleton standing on the ground are:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \mJ_w^{\tr}(\vq) \vf_w = \vtau + \mJ_o^{\tr}(\vq) \vf_o
\label{eq:dynamics}
\end{equation}
where $\vq$ are the joint angles of
the exoskeleton, $\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
velocities and accelerations, $\mM$ is the inertia matrix, $\mC()$ are the Coriolis and
centripetal forces, $\mG()$ are the gravitational forces, $\mF()$ are the
friction forces, $\mJ_w^{\tr}(\vq) \vf_w$ (Jacobian matrix multiplied with the
contact forces)
are the exoskeleton-world
contact forces 
expressed as exoskeleton joint torques, 
$\vtau$ are the exoskeleton
joint torques,
and $\mJ_o^{\tr}(\vq) \vf_o$
are the operator-exoskelton forces expressed
as exoskeleton joint torques.

We do not describe how to derive this equation or identify appropriate parameters
for an exoskeleton. We can provide this information if requested.
We are assuming that the equations have been derived taking into account constraints
such as one or two feet on the ground. Until we discuss ``floating body'' dynamics
in Section~\ref{sec:float}, we are assuming contacts do not slip and are not broken.
We are assuming all degrees of freedom (joints) are actuated until 
Section~\ref{sec:float}.

To understand these equations better, it is useful to have all Jacobians equal
to the identity matrix. This means the operator is directly applying torques
at the exoskeleton joints ($\vtau_o$) (a realistic assumption), 
and so is the world ($\vtau_w$) (a very unrealistic assumption). In this case:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \vtau_w = \vtau + \vtau_o
\end{equation}

Note that we are ignoring the human operator dynamics and operator muscle and tissue
forces resulting in operator joint torques. Intent prediction will model these
effects [intent.pdf].

We are also ignoring actuator dynamics, which we can handle by augmenting 
the state vector and adding additional equations.

@@

One control option is to use only passive controls, and no active controls
or actuators.
Such a system would rely on counterbalancing, mechanical springs, air springs, etc.
to reduce forces. 
An operator would rely on mechanical backdrivability to drive
exokeleton to do tasks.
In this case, 
$\vtau$ is generated by passive devices such as springs or dampers.

@@

Another control option is open loop control, in which actuator commands are
constant, or a function of time or phase of a behavior or task
(at least on the small time scale).
Commands or behaviors may be selected at a relatively coarse time interval
by the operator or by the exoskeleton controller.
This approach typically 
uses position and velocity control in either joint or
Cartesian coordinates
with either pre-generated references, selected targets,
or online generated trajectories to move to targets or perform tasks.
The operator relies on mechanical backdrivability 
to correct exoskeleton behavior.
In this case, it is common to generate the exoskeleton actuator commands using
a time or phase index:
$\vtau = \vtau_{ff}(t)$ where $\vtau_{ff}$ is a time
dependent feedforward torque vector.

@@

Active gravity compensation is similar 
to counterbalancing or physical gravity compensation, except it is done
with exoskeleton actuators to cancel the exoskeleton weight (but not inertia).
\begin{equation}
\vtau = \hat{\mG}(\hat{\vq})
\end{equation}
where $\vtau$ are the exoskeleton joint torques, 
$\hat{\mG}()$ are the gravitational forces calculated using an imperfect model, 
and $\hat{\vq}$ is the estimated configuration of the exoskeleton,
The operator relies on mechanical backdrivability to drive
the exokeleton to do tasks.

@@

A simple approximation for a lower body exoskeleton with a heavy payload at the
torso is:
\begin{equation}
\vtau \approx \hat{\mJ}^{\tr}(\hat{\vq}) (\hat{m} \hat{\vg})
\end{equation}
where $\hat{\mJ}$ is the estimated Jacobian matrix,
$\hat{m}$ is the estimated
weight of the exoskeleton (which could include the operator),
and $\hat{\vg}$ is the gravity vector (it is estimated due to possible orientation
errors in the controller's estimate of vertical).
All weight is assumed to be concentrated at the center of mass of the full system.

@@

Friction can also be compensated.
\begin{equation}
\vtau = \hat{\mF}(\hat{\vq},\hat{\dot{\vq}}) 
\end{equation}
are the estimated friction torques at the estimated exoskeleton configuration
and velocity.
Note that these may be load dependent, and may require additional state to
correctly handle stiction and hysteresis effects.

@@

The goal here is to map from desired motion (in this case desired
accelerations) to exoskeleton actuator commands (and desired operator
and world contact forces).
The exoskeleton actuator torques
attempt to match the torques caused by inertial, Coriolis, centripetal,
gravitational, and frictional forces.
\begin{equation}
\vtau = \vtau_{\invdyn} 
= \hat{\mM}(\hat{\vq}) \ddot{\vq}_d
+ \hat{\mC}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mG}(\hat{\vq})
+ \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})
\end{equation}
If in addition we want to cancel exoskeleton-world contact forces,
we add another term:
\begin{equation}
\vtau = \vtau_{\invdynw} = \vtau_{\invdyn} 
+ \hat{\mJ}_w^{\tr}(\hat{\vq}) \hat{\vf}_w
\end{equation}
If there are actuator dynamics or other dynamics, we can include those
dynamics as well.

These force/torque estimates rely on estimated dynamic models and their parameters.
If the models and estimated state (position and velocity) are perfect, we get
perfect cancellation of most of the dynamics:
\begin{equation}
\mM(\vq) (\ddot{\vq} - \ddot{\vq}_d) = \mJ_o^{\tr}(\vq) \vf_o
\end{equation}
If $\ddot{\vq}_d$ is set to zero, the operator directly drives the acceleration
of the exoskeleton with an effective inertia of $(\mM^{-1}(\vq) \mJ_o^{\tr}(\vq))^{-1}$
\begin{equation}
\ddot{\vq} = \mM^{-1}(\vq) \mJ_o^{\tr}(\vq) \vf_o
\label{eq:cancel}
\end{equation}
$\ddot{\vq}_d$ can be used to drive the exoskeleton along trajectories, and
the operator can directly modify the trajectory.

Simplifying equation~\ref{eq:cancel} by 
assuming the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM (\vq) \ddot{\vq} = \vtau_o
\end{equation}
The operator sees the true inertia of the exoskeleton, but no gravitational
loads and no other dynamics (Coriolis, centripetal, or frictional).

Note that we cannot change the apparent inertia of the exoskeleton without
some form of acceleration or force feedback.
With acceleration feedback we can make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding acceleration feedback:
\begin{equation}
\vtau = \vtau_{\invdynw} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
\label{eq:change-inertia1}
\end{equation}
sets the apparent inertia to $\mM_d$ if the dynamic models are perfect, and
state estimation produces not only perfect positions and velocities but
also perfect acceleration estimates. In the case 
where the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_o
\end{equation}
In robotics this is usually considered a bad idea if $\mM_d$ is smaller (a complex
issue for a matrix) than $\mM$ at any $\vq$. This type of control is vulnerable
to unmodeled dynamics. However, in an exoskeleton, since we have not tried to
cancel the human's dynamics, the human stabilizes a possible unstable exoskeleton
(if the straps are tight enough).

With force feedback we can also make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding force feedback:
\begin{equation}
\vtau = \vtau_{\invdynw} + \mK \hat{\vf}_o
\end{equation}
In the case where the force measurements are perfect and 
the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
(\mK + \mI)^{-1} \mM (\vq) \ddot{\vq} = \vtau_o
\end{equation}
By setting $\mK = \mM^{-1}_d (\vq) \mM (\vq) - \mI$ we can achieve
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_o
\end{equation}

@@

A key question is where does the desired acceleration $\ddot{\vq}_d$ come from?
This is discussed in the companion paper [intent.pdf].

@@

Here we make the exoskeleton appear as a locally linear impedance in joint
coordinates to the operator.
We can also make the exoskeleton appear as a locally linear impedance to the
operator in some other coordinate system, such as Cartesian coordinates.
A different goal is to make the exoskeleton appear as a locally linear impedance to
the outside world in some coordinate system.

To make the exoskeleton appear as a locally linear impedance in joint
coordinates to the operator, we add position and velocity feedback to $\vtau$:
\begin{equation}
\vtau = \vtau_{\invdynw} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
\label{eq:impedance}
\end{equation}
so if the dynamic models and state estimation are perfect we get this impedance:
\begin{equation}
\mM \ddot{\vq} - \mK_{\dot{\vq}} \dot{\vq} - \mK_{\vq} ( \vq - \vq_d ) = \vtau_o
\end{equation}
and the operator sees the desired impedance.

@@

We can implement operator force control 
if measurements of the operator contact forces with
the exoskeleton are available.
The exoskeleton moves as to create as little force on the operator as possible 
assuming the actuators are torque sources:
\begin{equation}
\vtau = \vtau_{\invdynw} + \mJ_o^{\tr}(\vq) {\vf_o}_d
+ \mJ_o^{\tr}(\vq) \mK_{\vf_o} ( \hat{\vf}_o - {\vf_o}_d )
\end{equation}
$\mK_{\vf_o}$ is a gain matrix, $\mJ_o$ is the Jacobian matrix for the operator
force $\vf_o$
Note that this does not require inverting the Jacobian matrix.
We will see that the position control version does invert the Jacobian matrix.

Applying this to force between the operator and the exoskeleton, and assuming
that the operator directly applies exoskeleton joint torques,
\begin{equation}
\vtau = \vtau_{\invdynw} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
\end{equation}
So if our dynamic models and state estimation are perfect, we get:
\begin{equation}
\ddot{\vq} = - \mM^{-1} (\vq) ({\vtau_o}_d + \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d ))
\end{equation}

One can add force damping terms, which sometimes help:
\begin{equation}
\vtau = \vtau_{\invdynw} + {\vtau_o}_d - \mK_{\vtau_o} ( \hat{\vtau}_o - {\vtau_o}_d )
- \mK_{\dot{\vtau}_o} \hat{\dot{\vtau}}_o
\end{equation}

Note that impedance control can also be implemented using force control
by setting:
\begin{equation}
{\vtau_o}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
\end{equation}
in addition to directly including the stiffness and damping terms in
equation~\ref{eq:impedance}.

@@

A control objective could be to have the exoskeleton to autonomously
apply a desired force.
One could add
\begin{equation}
\vtau = \vtau_{\invdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d
\end{equation}
to a gravity compensated control or to control including inverse dynamics.

>>>>

\newcommand{\ff}{{\mbox {\tiny ff}}}
\newcommand{\myo}{{\mbox {\tiny o}}}
\newcommand{\myx}{{\mbox {\tiny x}}}
\newcommand{\ox}{{\mbox {\tiny ox}}}
\newcommand{\ow}{{\mbox {\tiny ow}}}
\newcommand{\xw}{{\mbox {\tiny xw}}}
\newcommand{\xinvdyn}{{\mbox {\tiny x-ID}}}
\newcommand{\xinvdynw}{{\mbox {\tiny x-ID+Fw}}}
\newcommand{\oxinvdyn}{{\mbox {\tiny ox-ID}}}
\newcommand{\oxinvdynw}{{\mbox {\tiny ox-ID+Fw}}}

@@

\title{A Survey of Possible Exoskeleton Control Architectures and
Algorithms\\
(Draft 3.0)}

@@

It is useful to provide human programmers
with idealized abstractions of control of a single degree of
freedom. Here are some common alternatives that could be provided by low level control:
\begin{itemize}
\item
Position control. The joint moves to a specified position and stays there independent
of load.
\item
Velocity control. The joint moves at a specified velocity independent of load.
\item
Force or torque control. The joint provides a specified force or torque independent
of its motion.
\item
Impedance control: The user specifies a desired position (for a spring), and commands
a desired
stiffness, damping, and potentially a modified inertia or moment of inertia.
Modifying inertia is more dangerous than providing variable active stiffness
or damping.
Or (almost equivalently),
the user specifies a desired force or torque
which changes according to the position, velocity,
and acceleration of the joint.
\item
Valve control (for fluidic systems):
\begin{equation}
\vu = \mK_{\vq} (\vq - \vq_d) + \mK_{\dot{\vq}} ( \dot{\vq} - \dot{\vq}_d )
+ \mK_{\vtau} ( \vtau - \vtau_d )
+ \mK_{\dot{\vtau}} ( \dot{\vtau} - \dot{\vtau}_d ) + \vu_{\ff}
\end{equation}
where $\vu$ is a valve command vector (with a valve command for each actuator),
$\vq$ is the vector of joint angles, $\dot{\vq}$ are joint velocities,
$\vtau$ are actutor forces or joint torques, and $\dot{\vtau}$ is the rate of change
of the forces or torques (the time derivative of $\vtau$). 
The subscript $d$ indicates a desired value,
$\mK_{\vq}$ is a joint position gain matrix,
$\mK_{\dot{\vq}}$ is a joint velocity gain matrix, 
$\mK_{\vtau}$ is a joint force/torque gain matrix,
and $\mK_{\dot{\vtau}}$ is a joint force/torque derivative gain matrix.
$\vu_{\ff}$ are feedforward valve commands.
This formulation takes into account the most important hydraulic actuator dynamics,
and allows full state feedback
(including $\vtau$ as part of the state), pole placement, and linear quadratic
regulator (LQR) control of
a third order model of actuator plus rigid body dynamics.
If electric motor and power amplifier dynamics are taken into account in electric
motor systems, there would be an analogous equation for the commanded
motor voltage of
an analog power amplifier or
the control input of a PWM amplifier.
The $\mK_{\vtau} ( \vtau - \vtau_d )$ could be replaced or augmented
by a servo on piston differential pressure in a fluidic system or
a current servo in an electrical system.
\item
Online optimization of accelerations, actuator forces and torques, and contact forces
using quadratic programming. See Section \ref{sec:qp} for further explanation.
\end{itemize}

@@

PID or more complex nonlinear feedback control laws such as sliding mode control
are implemented by low level control. Note that a PID control law can be made
(smoothly) nonlinear by changing the PID gains on each servo tick.
Distributed low level control with limited communication typically controls each
joint or each set of joints (a limb, for example) independently.
Centralized low level control can support cross coupling of all joints,
and implement ``modern'' control schemes such as full state feedback,
pole placement, linear quadratic regulator (LQR) design, linear quadratic
Gaussian (LQG) design, and linear quadratic Gaussian loop transfer recovery
(LQG-LTR) design,

@@

\section{Operator and exoskeleton dynamics}

Exoskeletons are similar to robots, but there are several important differences.
There are two sets of contact forces: operator-exoskeleton (which we denote
with a subscript $\ox$ as in $\vf_{\ox}$) and exoskeleton-world (which we denote
with a subscript $\xw$ as in $\vf_{\xw}$). If the operator is tightly strapped or
rigidly held by other types of physical constraints, we can lump the operator
body parts with the exoskeleton parts which they are attached to in terms
of link inertias, locations of center of mass, and moments of inertia,
and just think of the whole system as one robot (we indicate this with
the subscript $\ox$ in dynamics equations.
We can also separate out the operator dynamics (indicated by
the subscript $\myo$) from the exoskeleton dynamics (indicated by
the subscript $\myx$) if this is more convenient
(we separate them out and ignore them in what immediately follows).
If the operator is loosely connected to the exoskeleton, this introduces a
huge modeling and control problem, and would also be very dangerous for the
operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).
We do not address the loosely connected case in this paper.

The dynamics of an operator standing on the ground (or on parts of the exoskeleton
that are on the ground):
\begin{equation}
\mM_{\myo}(\vq) \ddot{\vq} + \mC_{\myo}(\vq,\dot{\vq}) + \mG_{\myo}(\vq)
+ \mJ_{\ow}^{\tr}(\vq) \vf_{\ow} = \vtau_{\myo} - \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
\label{eq:odynamics}
\end{equation}
where $\vq$ are the operator joint angles,
$\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
joint velocities and accelerations,
$\mM_{\myo}$ is the operator inertia matrix, 
$\mC_{\myo}()$ are the operator Coriolis and
centripetal forces, 
$\mG_{\myo}()$ are the operator gravitational forces,
$\mJ_{\ow}^{\tr}(\vq) \vf_{\ow}$ (Jacobian matrix multiplied with the
contact forces) are the operator-world contact forces 
expressed as exoskeleton joint torques, 
$\vtau$ are the operator joint torques due to muscle activity and passive tissue
mechanics,
and 
$\mJ_{\ox}^{\tr}(\vq) \vf_{\ox}$
are the operator-exoskeleton forces expressed
as operator joint torques.
We assume the operator joint friction is negligable.

We assume the exoskeleton joints match some or all of the human joints.
The dynamics of such an exoskeleton standing on the ground or attached to
the human are:
\begin{equation}
\mM_{\myx}(\vq) \ddot{\vq} + \mC_{\myx}(\vq,\dot{\vq}) + \mG_{\myx}(\vq)
+ \mF_{\myx}(\vq,\dot{\vq})
+ \mJ_{\xw}^{\tr}(\vq) \vf_{\xw} = \vtau_{\myx} + \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
\label{eq:xdynamics}
\end{equation}
$\mM_{\myx}$ is the exoskeleton inertia matrix,
$\mC_{\myx}()$ are the exoskeleton Coriolis and centripetal forces,
$\mG_{\myx}()$ are the exoskeleton gravitational forces,
$\mF_{\myx}()$ are the exoskeleton friction forces,
$\mJ_{\xw}^{\tr}(\vq) \vf_{\xw}$ (Jacobian matrix multiplied with the
contact forces) are the exoskeleton-world
contact forces expressed as operator joint torques, 
$\vtau_{\myx}$ are the exoskeleton joint torques,
and 
$\mJ_{\ox}^{\tr}(\vq) \vf_{\ow}$
are the operator-exoskeleton forces expressed
as operator joint torques.

The dynamics of the combined system when the operator is tightly strapped in
or otherwise ``rigidly'' attached to the exoskeleton are:
\begin{equation}
\mM_{\ox}(\vq) \ddot{\vq} + \mC_{\ox}(\vq,\dot{\vq}) + \mG_{\ox}(\vq)
+ \mF_{\myx}(\vq,\dot{\vq})
+ \mJ_{\ow}^{\tr}(\vq) \vf_{\ow}
+ \mJ_{\xw}^{\tr}(\vq) \vf_{\xw} 
= \vtau_{\myo} + \vtau_{\myx}
\label{eq:oxdynamics}
\end{equation}
$\mM_{\ox}$ is the combined operator-exoskeleton inertia matrix,
$\mC_{\ox}()$ are the combined Coriolis and centripetal forces,
and $\mG_{\ox}()$ are the combined gravitational forces.
Note that the forces between the operator and the exoskeleton
do not appear explicitly in the
combined dynamics, but they can be computed using either equation~\ref{eq:odynamics}
or \ref{eq:xdynamics}.

We are ignoring actuator dynamics, which we can handle by augmenting 
the state vector and adding additional equations.

We do not describe how to derive these equations or identify appropriate 
model parameters.
We can provide this information if requested.
We are assuming that the equations have been derived taking into account constraints
such as one or two feet on the ground. Until we discuss ``floating body'' dynamics
in Section~\ref{sec:float}, we are assuming contacts do not slip and are not broken.
Therefore foot touchdown and liftoff require special handling by controllers
using these dynamics, which we will not describe in this paper.
We are assuming all degrees of freedom (joints) are actuated by the operator,
but not necessarily by the exoskeleton, until we get to 
``floating body'' dynamics (Section~\ref{sec:float}).

To understand these equations better, it is useful to have all Jacobians equal
to the identity matrix. This means the operator is directly applying torques
at the exoskeleton joints ($\vtau_{\ox}$) (a realistic assumption), 
and so is the world ($\vtau_{\ow}$ and $\vtau_{\xw}$)
(a very unrealistic assumption). In this case,
the operator dynamics are:
\begin{equation}
\mM_{\myo}(\vq) \ddot{\vq} + \mC_{\myo}(\vq,\dot{\vq}) + \mG_{\myo}(\vq)
+ \vtau_{\ow} = \vtau_{\myo} - \vtau_{\ox}
\end{equation}
the exoskeleton dynamics are:
\begin{equation}
\mM_{\myx}(\vq) \ddot{\vq} + \mC_{\myx}(\vq,\dot{\vq}) + \mG_{\myx}(\vq)
+ \mF_{\myx}(\vq,\dot{\vq})
+ \vtau_{\xw} = \vtau_{\myx} + \vtau_{\ox}
\end{equation}
and the combined dynamics are:
\begin{equation}
\mM_{\ox}(\vq) \ddot{\vq} + \mC_{\ox}(\vq,\dot{\vq}) + \mG_{\ox}(\vq)
+ \mF_{\myx}(\vq,\dot{\vq})
+ \vtau_{\ow}
+ \vtau_{\xw} 
= \vtau_{\myo} + \vtau_{\myx}
\label{eq:oxdynamics}
\end{equation}

\section{What do we want from intent prediction?}

For many of the following control systems, no operator intent estimation or
prediction is needed.
When we get to inverse dynamics, we need to know a desired acceleration
$\ddot{\vq}_d$.
In addition,
operator force 
estimates ($\hat{\vf}_{\myo}$ or $\hat{\vtau}_{\myo}$) or
operator-exoskeleton force 
estimates ($\hat{\vf}_{\ox}$ or $\hat{\vtau}_{\ox}$) are useful.
For online optimization, the more we know about the future, the better online
optimization can perform. We discuss this in Section~\ref{sec:predict}.

@@

One control option is to use only passive controls on a joint or on
all joints, and no active controls
or actuators.
Such a system would rely on counterbalancing, mechanical springs, air springs, etc.
to reduce forces. 
An operator would rely on mechanical backdrivability to drive
exokeleton to do tasks.
These passive devices have no controls, and can be included in the dynamics equations.
In this case, 
$\vtau_{\myx}$ is zero for any passive degree of freedom.

@@

Another control option is open loop control, in which actuator commands are
constant, or a function of time or phase of a behavior or task
(at least on the small time scale).
Commands or behaviors may be selected at a relatively coarse time interval
by the operator or by the exoskeleton controller.
This approach typically 
uses impedance control in either joint or
Cartesian coordinates
with either pre-generated references, selected targets,
or online generated trajectories to move to targets or perform tasks.
The operator relies on mechanical backdrivability 
to correct exoskeleton behavior.
In this case, it is common to generate the exoskeleton actuator commands using
a time or phase index:
$\vtau_{\myx} = \vtau_{\ff}(t)$ where $\vtau_{\ff}$ is a time or behavioral phase
dependent feedforward torque vector.

@@

Active exoskeleton gravity compensation is similar 
to counterbalancing or physical gravity compensation, except it is done
with exoskeleton actuators to cancel the exoskeleton weight (but not inertia):
\begin{equation}
\vtau_{\myx} = \hat{\mG}_{\myx}(\hat{\vq})
\end{equation}
where $\vtau_{\myx}$ are the exoskeleton joint torques, 
$\hat{\mG}_{\myx}()$ are the exoskeleton
gravitational forces calculated using an imperfect model, 
and $\hat{\vq}$ is the estimated configuration of the operator (and exoskeleton),
The operator relies on mechanical backdrivability to drive
the exokeleton to do tasks.
The weight of the operator can also be supported by the exoskeleton in
full gravity compensation:
\begin{equation}
\vtau_{\myx} = \hat{\mG}_{\ox}(\hat{\vq})
\end{equation}
where $\hat{\mG}_{\ox}()$ are the combined operator and exoskeleton
gravitational forces calculated using an imperfect model,

@@

A simple approximation for a lower body exoskeleton with a heavy payload at the
torso is:
\begin{equation}
\vtau_{\myx} \approx \hat{\mJ}^{\tr}(\hat{\vq}) (\hat{m} \hat{\vg})
\end{equation}
where $\hat{\mJ}$ is the estimated Jacobian matrix for the center of mass,
$\hat{m}$ is the estimated
weight of the exoskeleton (which could include the operator),
and $\hat{\vg}$ is the gravity vector (it is estimated due to possible orientation
errors in the controller's estimate of vertical).
All weight is assumed to be concentrated at the center of mass of the full system.

@@

Friction can also be compensated.
\begin{equation}
\vtau_{\myx} = \hat{\mF}_{\myx}(\hat{\vq},\hat{\dot{\vq}}) 
\end{equation}
are the estimated friction torques at the estimated exoskeleton configuration
and velocity.
Note that these may be load dependent, and may require additional state to
correctly handle stiction and hysteresis effects.

@@

The goal here is to map from desired motion (in this case desired
accelerations) to exoskeleton actuator commands (and desired operator
and world contact forces).
The exoskeleton actuator torques
attempt to match the torques caused by exoskeleton inertial, Coriolis, centripetal,
gravitational, and frictional forces.
\begin{equation}
\vtau_{\myx} = \vtau_{\xinvdyn} 
= \hat{\mM}_{\myx}(\hat{\vq}) \ddot{\vq}_d
+ \hat{\mC}_{\myx}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mG}_{\myx}(\hat{\vq})
+ \hat{\mF}_{\myx}(\hat{\vq},\hat{\dot{\vq}})
\end{equation}
If in addition we want to cancel exoskeleton-world contact forces,
we add another term:
\begin{equation}
\vtau_{\myx} = \vtau_{\xinvdynw} = \vtau_{\xinvdyn} 
+ \hat{\mJ}_{\xw}^{\tr}(\hat{\vq}) \hat{\vf}_{\xw}
\end{equation}
If there are actuator dynamics or other dynamics, we can include those
dynamics as well.

These force/torque estimates rely on estimated dynamic models and their parameters.
If the models and estimated state (position and velocity) are perfect, we get
perfect cancellation of most of the dynamics:
\begin{equation}
\mM_{\myx}(\vq) (\ddot{\vq} - \ddot{\vq}_d) = \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
\end{equation}
If $\ddot{\vq}_d$ is set to zero, the operator directly drives the acceleration
of the exoskeleton with an effective inertia of 
$(\mM^{-1}_{\myx}(\vq) \mJ_{\ox}^{\tr}(\vq))^{-1}$
\begin{equation}
\ddot{\vq} = \mM^{-1}_{\myx}(\vq) \mJ_{\ox}^{\tr}(\vq) \vf_{\ox}
\label{eq:cancel}
\end{equation}
$\ddot{\vq}_d$ can be used to drive the exoskeleton along trajectories, and
the operator can directly modify the current trajectory.

Simplifying equation~\ref{eq:cancel} by 
assuming the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM_{\myx} (\vq) \ddot{\vq} = \vtau_{\ox}
\end{equation}
The operator sees the true inertia of the exoskeleton, but no gravitational
loads and no other dynamics (Coriolis, centripetal, or frictional).

Note that we cannot change the apparent inertia of the exoskeleton without
some form of acceleration or force feedback.
With acceleration feedback we can make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding acceleration feedback:
\begin{equation}
\vtau_{\myx} = \vtau_{\xinvdynw} - (\mM_d(\hat{\vq}) - \hat{\mM}_{\myx}(\hat{\vq})) \hat{\ddot{\vq}}
\label{eq:change-inertia1}
\end{equation}
sets the apparent exoskeleton inertia to $\mM_d$ if the dynamic models are perfect, and
state estimation produces not only perfect positions and velocities but
also perfect acceleration estimates. In the case 
where the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_{\ox}
\end{equation}
In robotics this is usually considered a bad idea if $\mM_d$ is smaller (a complex
issue for a matrix) than $\mM_{\myx}$ at any $\vq$. This type of control is vulnerable
to unmodeled dynamics. However, in an exoskeleton, since we have not tried to
cancel the human's dynamics, the human stabilizes a possibly unstable exoskeleton
(if the straps are tight enough).

With force feedback we can also make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding force feedback:
\begin{equation}
\vtau_{\myx} = \vtau_{\xinvdynw} + \mK \hat{\vf}_{\ox}
\end{equation}
In the case where the force measurements are perfect and 
the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
(\mK + \mI)^{-1} \mM_{\myx} (\vq) \ddot{\vq} = \vtau_{\ox}
\end{equation}
By setting $\mK = \mM^{-1}_d (\vq) \mM_{\myx} (\vq) - \mI$ we can achieve
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_{\ox}
\end{equation}

@@

A key question is where does the desired acceleration $\ddot{\vq}_d$ come from?
This is discussed in a companion paper [intent.pdf].

@@

Here we make the exoskeleton appear as a locally linear impedance in joint
coordinates to the operator.
We can also make the exoskeleton appear as a locally linear impedance to the
operator in some other coordinate system, such as Cartesian coordinates.
A different goal is to make the exoskeleton or combined operator and
exoskeleton appear as a locally linear impedance to
the outside world in some coordinate system.

To make the exoskeleton appear as a locally linear impedance in joint
coordinates to the operator, we add position and velocity feedback to $\vtau_{\myx}$:
\begin{equation}
\vtau_{\myx} = \vtau_{\xinvdynw} - \mK_{\vq} ( \hat{\vq} - \vq_d ) - \mK_{\dot{\vq}} \hat{\dot{\vq}}
\label{eq:impedance}
\end{equation}
so if the dynamic models and state estimation are perfect we get this impedance:
\begin{equation}
\mM_{\myx} \ddot{\vq} + \mK_{\dot{\vq}} \dot{\vq} + \mK_{\vq} ( \vq - \vq_d ) = \vtau_{\ox}
\end{equation}
and the operator sees the desired impedance.

@@

We can implement operator force control 
if measurements of the operator contact forces with
the exoskeleton are available.
The exoskeleton moves as to create as little force on the operator as possible 
assuming the actuators are torque sources:
\begin{equation}
\vtau_{\myx} = \vtau_{\xinvdynw} + \mJ_{\ox}^{\tr}(\vq) {\vf_{\ox}}_d
+ \mJ_{\ox}^{\tr}(\vq) \mK_{\vf_{\ox}} ( \hat{\vf}_{\ox} - {\vf_{\ox}}_d )
\end{equation}
$\mK_{\vf_{\ox}}$ is a gain matrix, $\mJ_{\ox}$ is the Jacobian matrix for the operator
force $\vf_{\ox}$
Note that this does not require inverting the Jacobian matrix.
We will see that the position control version does invert the Jacobian matrix.

Applying this to force between the operator and the exoskeleton, and assuming
that the operator directly applies exoskeleton joint torques,
\begin{equation}
\vtau_{\myx} = \vtau_{\xinvdynw} + {\vtau_{\ox}}_d - \mK_{\vtau_{\ox}} ( \hat{\vtau}_{\ox} - {\vtau_{\ox}}_d )
\end{equation}
So if our dynamic models and state estimation are perfect, we get:
\begin{equation}
\ddot{\vq} = - \mM^{-1}_{\myx} (\vq) ({\vtau_{\ox}}_d + \mK_{\vtau_{\ox}} ( \hat{\vtau}_{\ox} - {\vtau_{\ox}}_d ))
\end{equation}

One can add force damping terms, which sometimes help:
\begin{equation}
\vtau = \vtau_{\xinvdynw} + {\vtau_{\ox}}_d - \mK_{\vtau_{\ox}} ( \hat{\vtau}_{\ox} - {\vtau_{\ox}}_d )
- \mK_{\dot{\vtau}_{\ox}} \hat{\dot{\vtau}}_{\ox}
\end{equation}

Note that impedance control can also be implemented using force control
by setting:
\begin{equation}
{\vtau_{\ox}}_d = \mK_{\vq}( \hat{\vq} - \vq_d ) + \mK_{\dot{\vq}} \hat{\dot{\vq}}
\end{equation}
in addition to directly including the stiffness and damping terms in
equation~\ref{eq:impedance}.

@@

A control objective could be to have the exoskeleton to autonomously
apply a desired force.
One could add
\begin{equation}
\vtau = \vtau_{\xinvdyn} + \mJ_w^{\tr}(\vq) {\vf_w}_d
\end{equation}
to a gravity compensated control or to control including inverse dynamics.
